---
title: "Exploration of temporal dynamics"
author: "Scott Forrest"
date: "`r Sys.Date()`"
execute: 
  cache: false
bibliography: paperpile.bib
toc: true
number-sections: false
format: 
  html:
    # self-contained: true
    code-fold: show
    code-tools: true
    df-print: paged
    code-line-numbers: true
    code-overflow: scroll
    fig-format: png
    fig-dpi: 300
  pdf:
    geometry: 
      - top=30mm
      - left=30mm
prefer-html: true
editor:
  source
abstract: |
  In this script we demonstrate different approaches to including covariates in an SSF model. 
---

# Load required packages

```{r}
#| warning=FALSE

library(tidyverse)
packages <- c("amt", "sf", "terra", "RColorBrewer", "leaflet")
walk(packages, require, character.only = T)

```

# Import data and clean

```{r}

buffalo_data <- read_csv("data/buffalo.csv") 

# remove individuals that have poor data quality or less than about 3 months of data. 
# The "2014.GPS_COMPACT copy.csv" string is a duplicate of ID 2024, so we exclude it
buffalo_data <- buffalo_data %>% filter(!node %in% c("2014.GPS_COMPACT copy.csv", 
                                           2029, 2043, 2265, 2284, 2346))

buffalo_data <- buffalo_data %>%  
  group_by(node) %>% 
  arrange(DateTime, .by_group = T) %>% 
  distinct(DateTime, .keep_all = T) %>% 
  arrange(node) %>% 
  mutate(ID = node)

buffalo_clean <- buffalo_data[, c(12, 2, 4, 3)]
colnames(buffalo_clean) <- c("id", "time", "lon", "lat")
attr(buffalo_clean$time, "tzone") <- "Australia/Queensland"
head(buffalo_clean)
tz(buffalo_clean$time)

buffalo_ids <- unique(buffalo_clean$id)

```

# Create a step object

Use the `amt` package to create a trajectory object from the cleaned data.

```{r}
#| label: make_track
#| code-summary: "Create a trajectory object"

buffalo_all <- buffalo_clean %>% mk_track(id = id,
                                           lon,
                                           lat, 
                                           time, 
                                           all_cols = T,
                                           crs = 4326) %>% 
  transform_coords(crs_to = 3112, crs_from = 4326) # Transformation to GDA94 / 
# Geoscience Australia Lambert (https://epsg.io/3112)

```

## Plot the data spatially

```{r}

buffalo_all %>%
  ggplot(aes(x = x_, y = y_, colour = id)) +
  geom_point(alpha = 0.5, size = 0.1) + 
  coord_fixed() +
  scale_x_continuous("Easting (m)") +
  scale_y_continuous("Northing (m)") +
  scale_colour_viridis_d() +
  theme_classic() +
  theme(legend.position = "right") 

# ggsave("outputs/data_prep/buffalo_djelk_map.png",
#        width = 150, height = 150, units = "mm",  dpi = 600)

```

## Show data on a satellite basemap

If viewing as a PDF this will be an automated screenshot. It viewing the HTML file this should be interactive.

```{r}
#| message: false
#| warning: false

# Use the original longitude/latitude data for mapping
leaflet(data = buffalo_clean) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addCircleMarkers(
    lng = ~lon, lat = ~lat,
    color = "red", radius = 0.5, opacity = 0.5,
    popup = ~paste("ID:", id, "<br>", "Time:", time)
  ) %>%
  addLayersControl(
    baseGroups = c("Satellite"),
    options = layersControlOptions(collapsed = TRUE)
  )

```

## Pick out a single individual

```{r}

which_buffalo <- "2022" # select a single buffalo ID

buffalo_id <- buffalo_all %>% filter(id == which_buffalo)

buffalo_id %>%
  ggplot(aes(x = x_, y = y_, colour = t_)) +
  geom_point(alpha = 0.5, size = 0.1) + 
  coord_fixed() +
  scale_x_continuous("Easting (m)") +
  scale_y_continuous("Northing (m)") +
  theme_classic()

```

## Create steps

We will just use a steps object here for now

```{r}

buffalo_id_steps <- buffalo_id %>% 
  steps()

head(buffalo_id_steps)

```

# Import spatial covariates

## NDVI

Although NDVI changes over time, and we have access to monthly layers, we will just select a single month here.

```{r}

ndvi <- rast("mapping/ndvi_aug_2018.tif")
plot(ndvi, main = "NDVI August 2018")
points(buffalo_id$x_, buffalo_id$y_, col = "red", pch = 16, cex = 0.5)
ndvi

```

## Other covariates

```{r}

canopy <- rast("mapping/canopy_cover.tif")
herby <- rast("mapping/veg_herby.tif")
slope <- rast("mapping/slope_raster.tif")

spatial_covs <- c(ndvi, canopy, herby, slope)
names(spatial_covs) <- c("ndvi", "canopy", "herby", "slope")
plot(spatial_covs)

```

## Extract covariate values at the end of each step

```{r}

buffalo_id_steps <- buffalo_id_steps %>% 
  extract_covariates(covariates = spatial_covs, where = "end")

head(buffalo_id_steps)

```

## Prepare data for extracting covariate information

```{r}

buffalo_id_steps <- buffalo_id_steps %>% 
  mutate(t1_ = lubridate::with_tz(buffalo_id_steps$t1_, tzone = "Australia/Darwin"),
         t2_ = lubridate::with_tz(buffalo_id_steps$t2_, tzone = "Australia/Darwin"))

buffalo_id_steps <- buffalo_id_steps %>%
  mutate(x1 = x1_, x2 = x2_, 
         y1 = y1_, y2 = y2_, 
         t1 = t1_, 
         t1_rounded = round_date(t1_, "hour"), 
         hour_t1 = hour(t1_rounded),
         t2 = t2_, 
         t2_rounded = round_date(t2_, "hour"), 
         hour_t2 = hour(t2_rounded),
         hour_t2 = ifelse(hour_t2 == 0, 24, hour_t2),
         yday = yday(t1_),
         year = year(t1_), 
         month = month(t1_),
         sl = sl_, 
         log_sl = log(sl_), 
         ta = ta_, 
         cos_ta = cos(ta_),
         canopy_01 = canopy/100)

head(buffalo_id_steps) 

```

# Hourly movement behaviour and selection of covariates

Here we bin the trajectories into the hours of the day, and calculate the mean, and quantiles for the step lengths and four habitat covariates. This is a similar approach to in @Forrest2024-gc and @Forrest2025-bn, where we used this approach to assess temporal dynamics as an exploratory step, but also as a validation step by assessing whether the simualted trajectories had also learned the temporally dynamic behaviour.

## Observed buffalo data

```{r}

buffalo_hourly_habitat <- 
  buffalo_id_steps %>% dplyr::group_by(hour_t2) %>% 
  summarise(n = n(),

            # step lengths
            step_length_mean = mean(sl_),
            step_length_q025 = quantile(sl_, 0.025),
            step_length_q25 = quantile(sl_, 0.25),
            step_length_median = quantile(sl_, 0.5),
            step_length_q75 = quantile(sl_, 0.75),
            step_length_q975 = quantile(sl_, 0.975),

            # ndvi
            ndvi_mean = mean(ndvi),
            ndvi_q025 = quantile(ndvi, 0.025),
            ndvi_q25 = quantile(ndvi, 0.25),
            ndvi_median = median(ndvi),
            ndvi_q75 = quantile(ndvi, 0.75),
            ndvi_q975 = quantile(ndvi, 0.975),

            # herby
            herby_mean = mean(herby),
            herby_q025 = quantile(herby, 0.025),
            herby_q25 = quantile(herby, 0.25),
            herby_median = median(herby),
            herby_q75 = quantile(herby, 0.75),
            herby_q975 = quantile(herby, 0.975),

            # canopy cover
            canopy_mean = mean(canopy_01),
            canopy_q025 = quantile(canopy_01, 0.025),
            canopy_q25 = quantile(canopy_01, 0.25),
            canopy_median = median(canopy_01),
            canopy_q75 = quantile(canopy_01, 0.75),
            canopy_q975 = quantile(canopy_01, 0.975),
            
            # slope
            slope_mean = mean(slope),
            slope_q025 = quantile(slope, 0.025),
            slope_q25 = quantile(slope, 0.25),
            slope_median = median(slope),
            slope_q75 = quantile(slope, 0.75),
            slope_q975 = quantile(slope, 0.975)
            
            ) %>% ungroup()

head(buffalo_hourly_habitat)

```

### Lengthen the dataframe

```{r}

buffalo_hourly_habitat_long <- buffalo_hourly_habitat %>%
  pivot_longer(cols = !c(hour_t2), names_to = "variable", values_to = "value") %>% 
  filter(!variable == "n")

head(buffalo_hourly_habitat_long)

```

### Set plotting parameters

```{r}

# set plotting parameters here that will change in each plot

# path
path_linewidth <- 0.5
path_alpha <- 0.1
path_95_alpha <- 1

# ribbon
ribbon_95_alpha <- 0.5
ribbon_50_alpha <- 0.25

```

### Loop over each variable to create a plot

As we have some categorical and binary variables, we use the **mean** for the line, rather than the median, and it explains why some of the ribbons look weird.

```{r}

# variables <- unique(buffalo_hourly_habitat_long$variable)

# variables to loop over
variables <- c("step_length", "ndvi", "herby", "canopy", "slope")

for(i in 1:length(variables)){
  
  # Filter data for the current variable
  var_data <- buffalo_hourly_habitat_long %>% 
    filter(str_detect(variable, variables[i]))
  
  # Extract specific quantiles for ribbons
  var_summary <- buffalo_hourly_habitat %>% 
    select(hour_t2, 
           mean = !!paste0(variables[i], "_mean"),
           q025 = !!paste0(variables[i], "_q025"), 
           q25 = !!paste0(variables[i], "_q25"),
           median = !!paste0(variables[i], "_median"),
           q75 = !!paste0(variables[i], "_q75"),
           q975 = !!paste0(variables[i], "_q975"))
  
  plot <- ggplot(data = var_summary) +
    
    # ribbons
    # 95% ribbon
    # geom_ribbon(aes(x = hour_t2, 
    #                 ymin = q025,
    #                 ymax = q975),
    #             alpha = ribbon_95_alpha, fill = "grey70") +
    
    # 50% ribbon
    geom_ribbon(aes(x = hour_t2, 
                    ymin = q25,
                    ymax = q75),
                alpha = ribbon_50_alpha, fill = "grey50") +
    
    # lines
    geom_line(aes(x = hour_t2, y = mean), size = 1) +
    geom_point(aes(x = hour_t2, y = mean), size = 2) +
    
    scale_x_continuous(breaks = seq(0, 24, by = 2)) +
    labs(x = "Hour of the day",
         y = variables[i],
         title = paste0("Hourly ", variables[i])) +
    theme_bw()
  
  print(plot)

  # ggsave(paste0("outputs/exploratory_temporal_dynamics/buffalo_hourly_", variables[i], ".png"),
  #        plot = plot,
  #        width = 150, height = 100, units = "mm", dpi = 600)
  
}

```

### References

::: {#refs}
:::
