---
title: "Assessing environmental extrapolation"
author: "Scott Forrest"
date: "`r Sys.Date()`"
execute: 
  cache: false
bibliography: paperpile.bib
toc: true
number-sections: false
format: 
  html:
    # self-contained: true
    code-fold: show
    code-tools: true
    df-print: paged
    code-line-numbers: true
    code-overflow: scroll
    fig-format: png
    fig-dpi: 300
  pdf:
    fig-format: png
    fig-dpi: 300
    geometry: 
      - top=30mm
      - left=30mm
editor:
  source
abstract: |
  Here we show how to assess extrapolation (in environmental space), using the the continuous Shape metric [@Velazco2024-xs]. 
---

# Load required packages

```{r}
#| warning=FALSE

# remotes::install_github("sjevelazco/flexsdm")

library(tidyverse)
packages <- c("amt", "sf", "terra", "tictoc", "beepr", "flexsdm", "tmap", "patchwork", "ggspatial")
walk(packages, require, character.only = T)

```

# Import data and clean

```{r}

buffalo_data <- read_csv("data/buffalo.csv") 

# remove individuals that have poor data quality or less than about 3 months of data. 
# The "2014.GPS_COMPACT copy.csv" string is a duplicate of ID 2024, so we exclude it
buffalo_data <- buffalo_data %>% filter(!node %in% c("2014.GPS_COMPACT copy.csv", 
                                           2029, 2043, 2265, 2284, 2346))

buffalo_data <- buffalo_data %>%  
  group_by(node) %>% 
  arrange(DateTime, .by_group = T) %>% 
  distinct(DateTime, .keep_all = T) %>% 
  arrange(node) %>% 
  mutate(ID = node)

buffalo_clean <- buffalo_data[, c(12, 2, 4, 3)]
colnames(buffalo_clean) <- c("id", "time", "lon", "lat")
attr(buffalo_clean$time, "tzone") <- "Australia/Queensland"
buffalo_clean$sex <- "f"
buffalo_clean$life_stage <- "adult"
head(buffalo_clean)
tz(buffalo_clean$time)

buffalo_ids <- unique(buffalo_clean$id)

write_csv(buffalo_clean, "data/buffalo_clean.csv")

```

# Create a step object 

Use the `amt` package to create a trajectory object from the cleaned data. 

```{r}
#| label: make_track
#| code-summary: "Create a trajectory object"

buffalo_all <- buffalo_clean %>% mk_track(id = id,
                                           lon,
                                           lat, 
                                           time, 
                                           all_cols = T,
                                           crs = 4326) %>% 
  transform_coords(crs_to = 3112, crs_from = 4326) # Transformation to GDA94 / 
# Geoscience Australia Lambert (https://epsg.io/3112)

```

## Plot the data spatially

```{r}

buffalo_all %>%
  ggplot(aes(x = x_, y = y_, colour = id)) +
  geom_point(alpha = 0.5, size = 0.1) + 
  coord_fixed() +
  scale_x_continuous("Easting (m)") +
  scale_y_continuous("Northing (m)") +
  scale_colour_viridis_d() +
  theme_classic() +
  theme(legend.position = "right") 

# ggsave("outputs/data_prep/buffalo_djelk_map.png",
#        width = 150, height = 150, units = "mm",  dpi = 600)

```

## Creating a step object

```{r}
#| label: steps
#| code-summary: "Create steps"

# nest the data by individual
buffalo_all_nested <- buffalo_all %>% arrange(id) %>% nest(data = -"id")

buffalo_all_nested_steps <- buffalo_all_nested %>%
  mutate(steps = map(data, function(x)
    x %>% track_resample(rate = hours(1), tolerance = minutes(10)) %>%
      steps()))

# unnest the data after creating 'steps' objects
buffalo_all_steps <- buffalo_all_nested_steps %>% 
  amt::select(id, steps) %>% 
  amt::unnest(cols = steps)

buffalo_all_steps <- buffalo_all_steps %>% 
  mutate(t2_rounded = round_date(t2_, "hour"), # round the time to the nearest hour
         hour_t2 = ifelse(hour(t2_rounded) == 0, 24, hour(t2_rounded))) # change the 0 hour to 24

head(buffalo_all_steps, 10)

```

## Fitting step length and turning angle distributions

Fitting exponential and von Mises distributions to the steps of ALL individuals (only one Gamma and one von Mises distribution for the whole population). This should be done when fitting a hierarchical model to update the 'population' parameters, but also makes it straightforward to update after model fitting to each individual separately.

```{r}
#| label: fit_distributions
#| code-summary: "Fit step length and turning angle distributions"

# fitting step length and turning angle distributions to all locations
gamma_dist <- fit_distr(buffalo_all_steps$sl_, "gamma")
vonmises_dist <- fit_distr(buffalo_all_steps$ta_, "vonmises")

# checking parameters - which can then be saved to update movement parameters 
# after fitting the step selection model
gamma_dist$params$shape
gamma_dist$params$scale
vonmises_dist$params$kappa
vonmises_dist$params$mu

```

For some reason, the `random_steps` function does not work when using the `bursted_steps_xyt` class. I'm not sure why (the error is `Error in bursts[[i]] : subscript out of bounds`), but it works when that class label is removed, and appears to sample random steps correctly. For taxa that have irregular fixes and many bursts, this may be worth exploring in more detail.

```{r}
#| label: random_steps
#| code-summary: "Create random steps"

tic()

buffalo_parametric_popn_GvM <- buffalo_all_steps %>% 
  random_steps(n_control = 10,
               sl_distr = gamma_dist,
               ta_distr = vonmises_dist) %>% 
  mutate(y = as.numeric(case_))

toc()

```

## Plotting the random step distributions

Spatially plot the used and random steps, with the used steps in red.

```{r}
#| label: plot_steps
#| code-summary: "Plot the random steps"

buffalo_parametric_popn_GvM %>% ggplot() +
  geom_point(data = . %>% filter(y == 0), aes(x = x2_, y = y2_), 
             colour = "black", size = 0.1, alpha = 0.5) +
  geom_point(data = . %>% filter(y == 1), aes(x = x2_, y = y2_), 
             colour = "red", size = 0.1, alpha = 0.5) +
  coord_equal() +
  theme_bw()

```

# Import spatial covariates

Although NDVI changes over time, and we have access to monthly layers, we will just select a single month here.

```{r}

# NDVI
# ndvi <- terra::rast("mapping/large_files/NDVI_2023_250m.tif")
# ndvi <- terra::project(ndvi, "EPSG:3112")
# writeRaster(ndvi, "mapping/large_files/NDVI_2023_250m_projected3112.tif", overwrite = TRUE)

ndvi <- terra::rast("mapping/large_files/NDVI_2023_250m_projected3112.tif")
plot(ndvi, main = "NDVI July 2023")
points(buffalo_all$x_, buffalo_all$y_, col = "red", pch = 16, cex = 0.5)

# elevation
# elev <- terra::rast("mapping/large_files/DEM_SRTM_250m.tif")
# elev <- terra::project(elev, "EPSG:3112")
# writeRaster(elev, "mapping/large_files/DEM_SRTM_250m_projected3112.tif", overwrite = TRUE)

elev <- terra::rast("mapping/large_files/DEM_SRTM_250m_projected3112.tif")
plot(elev, main = "Elevation")
points(buffalo_all$x_, buffalo_all$y_, col = "red", pch = 16, cex = 0.5)

# create slope layer
slope <- terra::terrain(elev, v = "slope", neighbors = 8)
plot(slope, main = "Slope")
points(buffalo_all$x_, buffalo_all$y_, col = "red", pch = 16, cex = 0.5)

```

## Sample values of the environmental covariates at the end of the steps.

```{r}
#| label: extract_covariates
#| code-summary: "Extract covariates at the end of the steps"

buffalo_parametric_popn_covs <- buffalo_parametric_popn_GvM %>% 
  
  extract_covariates(ndvi,
                     where = "end") %>%
  extract_covariates(elev,
                     where = "end") %>%
  extract_covariates(slope,
                     where = "end") %>% 
  
  mutate(y = as.numeric(case_),
         cos_ta_ = cos(ta_),
         log_sl_ = log(sl_))

```

# Assess extrapolation

```{r}

# uncomment to run the code - takes a while
# tic()
# 
# extrapolation_raster <- extra_eval(
#   training_data = buffalo_parametric_popn_covs,
#   pr_ab = "y",
#   projection_data = c(ndvi, elev, slope), 
#   metric = "mahalanobis",
#   univar_comb = FALSE,
#   aggreg_factor = 10
# )
# 
# beep(sound = 2)
# 
# toc()

```

## Write the raster 

```{r}

# terra::writeRaster(extrapolation_raster, "outputs/extrapolation_raster.tif", overwrite = TRUE)
extrapolation_raster <- terra::rast("outputs/extrapolation_raster.tif")

```

### Mask out NAs and plot

```{r}

extrapolation_raster <- terra::mask(extrapolation_raster, ndvi)
extrapolation_raster_wgs84 <- terra::project(extrapolation_raster, "EPSG:4326")
plot(extrapolation_raster_wgs84)

```

## Add Shape values back to the training dataframe

```{r}

buffalo_parametric_popn_covs <- buffalo_parametric_popn_covs %>% 
  extract_covariates(extrapolation_raster,
                     where = "end")

hist(buffalo_parametric_popn_covs$extrapolation, breaks = 30)

# what is the 99th percentile of the Shape values in the training data?
threshold_q99 <- quantile(buffalo_parametric_popn_covs$extrapolation, 0.99, na.rm = TRUE)
paste0("The 99th percentile of the Shape values in the training data is ", threshold_q99)

# what is the maximum Shape value in the training data?
threshold_max <- max(buffalo_parametric_popn_covs$extrapolation, na.rm = TRUE)
paste0("The maximum Shape value in the training data is ", threshold_max)

```

### Plot histogram of Shape values in the training data with the 99th percentile and maximum values indicated

```{r}

ggplot() +
  geom_histogram(data = buffalo_parametric_popn_covs, aes(x = extrapolation), 
                 fill = "orange", colour = "black", bins = 30) +
  geom_vline(xintercept = threshold_q99, color = "red", linetype = "dashed") +
  geom_vline(xintercept = threshold_max, color = "red", linetype = "dashed") +
  labs(x = "Extrapolation (Shape)", y = "Frequency") +
  theme_bw()

```

## Plot the extrapolation

### Create dataframe

```{r}

# prevent scientific notation in the legend
options(scipen=999)

extrapolation_raster_wgs84_agg <- terra::aggregate(extrapolation_raster_wgs84, fact = 4, fun = mean)
plot(extrapolation_raster_wgs84_agg)

extrapolation_raster_wgs84_agg_df <- as.data.frame(extrapolation_raster_wgs84_agg, xy = TRUE)

```

### Thin the training data

```{r}

buffalo_parametric_popn_covs_thin <- buffalo_parametric_popn_covs %>% dplyr::slice_sample(prop = 1)

# project to WGS84 for plotting with ggplot
xy_wgs84 <- buffalo_parametric_popn_covs_thin %>%
  st_as_sf(coords = c("x1_", "y1_"), crs = 3112) %>%
  st_transform(crs = 4326) %>%
  st_coordinates() %>%
  as.data.frame() %>%
  rename(x1_wgs84 = X, y1_wgs84 = Y)

buffalo_parametric_popn_covs_thin <- buffalo_parametric_popn_covs_thin %>%
  bind_cols(xy_wgs84) 

```

### Plot with ggplot

```{r}

raster_alpha <- 0.5
training_points_alpha <- 0.1

extrapolation_aus_plot <- ggplot(data = extrapolation_raster_wgs84_agg_df, aes(x = x, y = y, fill = extrapolation)) +
  
  geom_raster() +
    scale_fill_viridis_c(
    na.value = "transparent",
    transform = "log1p",
    breaks = c(1, 10, 100, 1000, 5000)
  ) +
  
  geom_contour(aes(z = extrapolation),
               breaks = c(threshold_max), color = "orange", linewidth = 0.2) +
  
  geom_point(data = buffalo_parametric_popn_covs_thin,
             aes(x = x1_wgs84, y = y1_wgs84),
             colour = "red", alpha = training_points_alpha, size = 0.01) +
  
  annotation_scale(location = "bl", width_hint = 0.2) +
  coord_sf(crs = 4326, default_crs = 4326, expand = FALSE) +
  labs(x = NULL, y = NULL, fill = "Extrapolation\n(Shape)") +
  theme_classic() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.95, 0.7)
    )

extrapolation_aus_plot

ggsave("outputs/extrapolation_raster_ggplot.png",
       width = 180, height = 60, units = "mm", scale = 1.5, dpi = 1000)

```

## Check distributions of covariates against Shape values

```{r}

cov_extrap_stack <- c(ndvi, elev, slope, extrapolation_raster)
cov_extrap_stack_df <- as.data.frame(cov_extrap_stack, xy = TRUE) %>% drop_na()

# thin to reduce computation and make plot look less messy
cov_extrap_stack_df_thin <- cov_extrap_stack_df %>% dplyr::slice_sample(prop = 0.001)

```

### Plot NDVI and elevation with Shape values

```{r}

ndvi_elev_plot <- ggplot() +
  
  geom_point(data = cov_extrap_stack_df_thin, 
             aes(x = NDVI, y = elevation, colour = extrapolation), alpha = raster_alpha) +
  
  geom_point(data = buffalo_parametric_popn_covs_thin,
             aes(x = NDVI, y = elevation),
             colour = "red", alpha = training_points_alpha) +
  
  scale_colour_viridis_c(
    transform = "log1p",
    breaks = c(1, 10, 100, 1000, 5000)
    ) +
  
  labs(x = "NDVI", y = "Elevation", colour = "Extrapolation\n(Shape)")  +
  theme_classic() +
  theme(legend.position = "none")

ndvi_elev_plot

```

### Plot NDVI and slope with Shape values

```{r}

ndvi_slope_plot <- ggplot() +
  
  geom_point(data = cov_extrap_stack_df_thin, 
             aes(x = NDVI, y = slope, colour = extrapolation), alpha = raster_alpha) +
  
  geom_point(data = buffalo_parametric_popn_covs_thin,
             aes(x = NDVI, y = slope),
             colour = "red", alpha = training_points_alpha) +
  
  scale_colour_viridis_c(
    transform = "log1p",
    breaks = c(1, 10, 100, 1000, 5000)
    ) +
  
  labs(x = "NDVI", y = "Slope", colour = "Extrapolation\n(Shape)") +
  theme_classic() +
  theme(legend.position = "none")

ndvi_slope_plot

```

### Plot elevation and slope with Shape values

```{r}

elev_slope_plot <- ggplot() +
  
  geom_point(data = cov_extrap_stack_df_thin, 
             aes(x = elevation, y = slope, colour = extrapolation), alpha = raster_alpha) +
  
  geom_point(data = buffalo_parametric_popn_covs_thin,
             aes(x = elevation, y = slope),
             colour = "red", alpha = training_points_alpha) +
  
  scale_colour_viridis_c(
    transform = "log1p",
    breaks = c(1, 10, 100, 1000, 5000)
    ) +
  
  labs(x = "Elevation", y = "Slope", colour = "Extrapolation\n(Shape)") +
  theme_classic() +
  theme(legend.position = "none")

elev_slope_plot

```

### Combine the plots

```{r}

extrapolation_aus_plot / 
  (ndvi_elev_plot | ndvi_slope_plot | elev_slope_plot) +
  plot_layout(heights = c(2, 1)) #+
  # plot_annotation(tag_levels = "A")

ggsave("outputs/extrapolation_aus_covs.png",
       width = 180, height = 110, units = "mm", scale = 1.5, dpi = 600)

```

# Create zoomed plot for earlier figure

## Crop spatial covariates

```{r}

# set the extent
zoom_extent_min_x <- min(buffalo_parametric_popn_covs_thin$x1_) - 20000
zoom_extent_max_x <- max(buffalo_parametric_popn_covs_thin$x1_) + 120000
zoom_extent_min_y <- min(buffalo_parametric_popn_covs_thin$y1_) - 50000
zoom_extent_max_y <- max(buffalo_parametric_popn_covs_thin$y1_) + 20000

# create extent
crop_extent <- terra::ext(
  zoom_extent_min_x,
  zoom_extent_max_x,
  zoom_extent_min_y,
  zoom_extent_max_y
)

ndvi_crop <- terra::crop(ndvi, crop_extent)
elev_crop <- terra::crop(elev, crop_extent)
slope_crop <- terra::crop(slope, crop_extent)

```

# Assess extrapolation

```{r}

# commented out here as it takes about 45 minutes to run
# tic()
# 
# extrapolation_raster_crop <- extra_eval(
#   training_data = buffalo_parametric_popn_covs,
#   pr_ab = "y",
#   projection_data = c(ndvi_crop, elev_crop, slope_crop),
#   metric = "mahalanobis",
#   univar_comb = FALSE,
#   aggreg_factor = 1
# )
# 
# beep(sound = 2)
# 
# toc()

# terra::writeRaster(extrapolation_raster_crop, "outputs/extrapolation_raster_crop.tif", overwrite = TRUE)
extrapolation_raster_crop <- terra::rast("outputs/extrapolation_raster_crop.tif")

```

```{r}

# mask and create dataframe for plotting
extrapolation_raster_crop <- terra::mask(extrapolation_raster_crop, ndvi_crop)
extrapolation_raster_cropped_df <- as.data.frame(extrapolation_raster_crop, xy = TRUE)

# plot with ggplot
ggplot(data = extrapolation_raster_cropped_df, aes(x = x, y = y, fill = extrapolation)) +
  
  geom_raster() +
  
  scale_fill_viridis_c(
    na.value = "transparent",
    transform = "log1p",
    breaks = c(1, 10, 100, 500)
  ) +
  
  geom_contour(aes(z = extrapolation),
               breaks = c(threshold_max), color = "orange", linewidth = 0.3) +
  
  geom_point(data = buffalo_parametric_popn_covs %>% filter(y == 0),
             aes(x = x2_, y = y2_),
             colour = "black", alpha = 0.25, size = 0.01) +
  
  geom_point(data = buffalo_parametric_popn_covs %>% filter(y == 1),
             aes(x = x1_, y = y1_),
             colour = "red", alpha = 0.25, size = 0.01) +
  
  annotation_scale(location = "bl", width_hint = 0.2) +
  coord_sf(crs = 3112, default_crs = 3112, expand = FALSE) +
  labs(x = NULL, y = NULL, fill = "Extrapolation\n(Shape)") +
  theme_classic() +
  theme(
    legend.position = "right"
    )

ggsave("outputs/extrapolation_raster_cropped_ggplot.png",
       width = 180, height = 90, units = "mm", scale = 1.5, dpi = 600)

```

# References

::: {#refs}
:::

