---
title: "SSF Formulations"
author: "Scott Forrest"
date: "`r Sys.Date()`"
execute: 
  cache: false
bibliography: paperpile.bib
toc: true
number-sections: false
format: 
  html:
    # self-contained: true
    code-fold: show
    code-tools: true
    df-print: paged
    code-line-numbers: true
    code-overflow: scroll
    fig-format: png
    fig-dpi: 300
  pdf:
    fig-format: png
    fig-dpi: 300
    geometry: 
      - top=30mm
      - left=30mm
editor:
  source
abstract: |
  In this script we demonstrate different approaches to including covariates in an SSF model. 
---

# Load required packages

```{r}
#| warning=FALSE

library(tidyverse)
packages <- c("amt", "sf", "terra", "RColorBrewer", "mgcv", "gratia")
walk(packages, require, character.only = T)

```

# Import data and clean

```{r}

buffalo_data <- read_csv("data/buffalo.csv") 

# remove individuals that have poor data quality or less than about 3 months of data. 
# The "2014.GPS_COMPACT copy.csv" string is a duplicate of ID 2024, so we exclude it
buffalo_data <- buffalo_data %>% filter(!node %in% c("2014.GPS_COMPACT copy.csv", 
                                           2029, 2043, 2265, 2284, 2346))

buffalo_data <- buffalo_data %>%  
  group_by(node) %>% 
  arrange(DateTime, .by_group = T) %>% 
  distinct(DateTime, .keep_all = T) %>% 
  arrange(node) %>% 
  mutate(ID = node)

buffalo_clean <- buffalo_data[, c(12, 2, 4, 3)]
colnames(buffalo_clean) <- c("id", "time", "lon", "lat")
attr(buffalo_clean$time, "tzone") <- "Australia/Queensland"
buffalo_clean$sex <- "f"
buffalo_clean$life_stage <- "adult"
head(buffalo_clean)
tz(buffalo_clean$time)

buffalo_ids <- unique(buffalo_clean$id)

write_csv(buffalo_clean, "data/buffalo_clean.csv")

```

# Create a step object 

Use the `amt` package to create a trajectory object from the cleaned data. 

```{r}
#| label: make_track
#| code-summary: "Create a trajectory object"

buffalo_all <- buffalo_clean %>% mk_track(id = id,
                                           lon,
                                           lat, 
                                           time, 
                                           all_cols = T,
                                           crs = 4326) %>% 
  transform_coords(crs_to = 3112, crs_from = 4326) # Transformation to GDA94 / 
# Geoscience Australia Lambert (https://epsg.io/3112)

```

## Plot the data spatially

```{r}

buffalo_all %>%
  ggplot(aes(x = x_, y = y_, colour = id)) +
  geom_point(alpha = 0.5, size = 0.1) + 
  coord_fixed() +
  scale_x_continuous("Easting (m)") +
  scale_y_continuous("Northing (m)") +
  scale_colour_viridis_d() +
  theme_classic() +
  theme(legend.position = "right") 

# ggsave("outputs/data_prep/buffalo_djelk_map.png",
#        width = 150, height = 150, units = "mm",  dpi = 600)

```

## Pick out a single individual

```{r}

which_buffalo <- "2158" # select a single buffalo ID

# create directory to save plots for each buffalo
plot_save_path <- paste0("outputs/buffalo_", which_buffalo)
dir.create(plot_save_path, showWarnings = FALSE)

buffalo_id <- buffalo_all %>% filter(id == which_buffalo)

buffalo_id %>%
  ggplot(aes(x = x_, y = y_, colour = t_)) +
  geom_path(alpha = 0.5) + 
  geom_point(alpha = 0.5, size = 0.1) + 
  coord_fixed() +
  scale_x_continuous("Easting (m)") +
  scale_y_continuous("Northing (m)") +
  theme_classic()

ggsave(paste0(plot_save_path, "/buffalo_id_", which_buffalo, "_map.png"),
       width = 150, height = 150, units = "mm",  dpi = 600)

```

# Import spatial covariate

Although NDVI changes over time, and we have access to monthly layers, we will just select a single month here.

```{r}

ndvi <- rast("mapping/ndvi_aug_2018.tif")
plot(ndvi, main = "NDVI August 2018")
points(buffalo_id$x_, buffalo_id$y_, col = "red", pch = 16, cex = 0.5)
ndvi

```

# Select local extent

We just want to look at an area centred on a single location, which might be combined with our movement probability surface to generate a next-step probability surface.

We'll centre the extent on an interesting looking part of the landscape, and then buffer it by 1500m (with a buffer of the cell resolution/2 to ensure there is a central cell), which contains most of the step lengths, and therefore most of the movement kernel.

```{r}

cell_resolution <- 25 # in metres

buffalo_single_point <- buffalo_id[7500,] # select one of the locations

buffer <- 1500 + (cell_resolution/2) # buffer in metres

window_extent <- ext(buffalo_single_point$x_ - buffer, 
                        buffalo_single_point$x_ + buffer, 
                        buffalo_single_point$y_ - buffer, 
                        buffalo_single_point$y_ + buffer)

ndvi_window <- crop(ndvi, window_extent)

# set the extent to 0 at the buffalo location
ndvi_window <- terra::shift(ndvi_window, 
                            dx = -buffalo_single_point$x_, 
                            dy = -buffalo_single_point$y_)

# plot the NDVI layer with the buffalo location as the centre point
plot(ndvi_window, 
     # main = paste0("NDVI"),
     col = brewer.pal(9, "Greens"), 
     plg=list(x="top", title="NDVI", tics="out", cex=.8))
points(x = 0, y = 0, col = "red")

# Save the plot
png(filename = paste0(plot_save_path, "/ndvi_local_", which_buffalo, ".png"), 
    width = 100, height = 100, units = "mm", res = 600)
plot(ndvi_window, 
     # main = paste0("NDVI"),
     col = brewer.pal(9, "Greens"), 
     plg=list(x="top", title="NDVI", tics="out", cex=.8))
points(x = 0, y = 0, col = "red")
dev.off()

```

## Sample random steps to fit models

```{r}

buffalo_id_steps <- buffalo_id %>% 
  steps()

# fitting step length and turning angle distributions to all locations
gamma_dist <- fit_distr(buffalo_id_steps$sl_, "gamma")
vonmises_dist <- fit_distr(buffalo_id_steps$ta_, "vonmises")

# movement parameters
gamma_dist$params$shape
gamma_dist$params$scale
vonmises_dist$params$kappa
vonmises_dist$params$mu

# sample random steps
buffalo_ssf_data <- buffalo_id_steps %>% 
  random_steps(n = 10, 
                sl_dist = gamma_dist, 
                ta_dist = vonmises_dist) %>%
  extract_covariates(ndvi)  %>%
  mutate(ndvi_sq = ndvi^2,
         log_sl = log(sl_),
         cos_ta = cos(ta_),
         times = 1) # create a dummy column for times (used in a Cox PH model, but not relevant to an SSF) that all contain the same value

```

# Different SSF formulations

Firstly, we want to pull out the range of NDVI values to plot the curves with.

```{r}

# histogram of NDVI values
hist(values(ndvi))

# 1% and 99% quantiles of NDVI values
ndvi_quantiles <- quantile(values(ndvi), probs = c(0.005, 0.995), na.rm = TRUE)
ndvi_quantiles

# plot across the range in the local window
ndvi_values <- seq(min(0, ndvi_quantiles[1]), 
                   ndvi_quantiles[2], 
                   length.out = 100)

pred_data <- data.frame(ndvi = ndvi_values, 
                        ndvi_sq = ndvi_values^2,
                        sl_ = 0,
                        log_sl = 0,
                        cos_ta = 0)

```

## Set RSS min and max values

This is hard-coded as the maximum rss value across the different formulations, to ensure the plotting extent is consistent across approaches.

```{r}

rss_min <- 0
rss_max <- 2.1

```

## Linear covariate

### Model formula

$y \sim ndvi$

### Fit model

```{r}

# fit an issf model
ssf_linear <- amt::fit_issf(case_ ~ 
                              ndvi + 
                              sl_ + log_sl + cos_ta +
                              strata(step_id_),
                            data = buffalo_ssf_data)

summary(ssf_linear)
AIC(ssf_linear)

# fit an ssf gam model with parametric terms
ssf_linear_gam <- mgcv::gam(cbind(times, step_id_) ~ 
                              ndvi +
                              sl_ + log_sl + cos_ta,
                            data = buffalo_ssf_data,
                            family = cox.ph,
                            weight = case_)

summary(ssf_linear_gam)
AIC(ssf_linear_gam)

```

```{r}

# Get predictions on log-scale with standard errors
predictions <- predict(ssf_linear_gam, 
                      newdata = pred_data, 
                      type = "link",  # log-scale
                      se.fit = TRUE)

# Extract log RSS and standard errors
log_rss <- predictions$fit
se_log_rss <- predictions$se.fit

# Calculate confidence intervals on log-scale
log_rss_lower <- log_rss - 1.96 * se_log_rss
log_rss_upper <- log_rss + 1.96 * se_log_rss

# Transform to RSS scale
rss <- exp(log_rss)
rss_lower <- exp(log_rss_lower)
rss_upper <- exp(log_rss_upper)

# Create your dataframe
ndvi_coef_df <- data.frame(
  ndvi_values = ndvi_values,
  log_rss = log_rss,
  log_rss_lower = log_rss_lower,
  log_rss_upper = log_rss_upper,
  rss = rss,
  rss_lower = rss_lower,
  rss_upper = rss_upper
)

# plot the response curve
ggplot(data = ndvi_coef_df) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey") +
  geom_ribbon(aes(x = ndvi_values, ymin = rss_lower, ymax = rss_upper), 
              fill = "grey70", alpha = 0.5) +
  geom_line(aes(x = ndvi_values, y = rss), 
            colour = "black", linewidth = 1) +
  scale_x_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  labs(x = "NDVI", y = "RSS") +
  theme_classic() 

```


### Response curve

```{r}

# extract the NDVI coefficient and its standard error
ndvi_coef <- ssf_linear$model$coefficients[1]
ndvi_coef_se <- sqrt(diag(ssf_linear$model$var))[1]  # SE of coefficient

# calculate the RSS for each NDVI value
ndvi_coef_df <- data.frame(
  ndvi_values = ndvi_values,
  log_rss = ndvi_values * ndvi_coef,
  rss = exp(ndvi_values * ndvi_coef),
  se_prediction = abs(ndvi_values) * ndvi_coef_se  # SE of prediction at each NDVI value
)

# plot the response curve
ggplot(data = ndvi_coef_df) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey") +
  geom_ribbon(aes(x = ndvi_values, ymin = exp(log_rss - 1.96 * se_prediction), ymax = exp(log_rss + 1.96 * se_prediction)), 
              fill = "grey70", alpha = 0.5) +
  geom_line(aes(x = ndvi_values, y = rss), 
            colour = "black", linewidth = 1) +
  geom_rug(data = buffalo_ssf_data %>% filter(case_ == T), 
           aes(x = ndvi), alpha = 0.25, sides="b") +
  scale_x_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  scale_y_continuous(limits = c(rss_min, rss_max)) +
  labs(x = "NDVI", y = "RSS") +
  theme_classic() 

ggsave(paste0(plot_save_path, "/ndvi_linear_response_id", which_buffalo, ".png"),
       width = 120, height = 100, units = "mm", dpi = 600)

```

### Plot habitat selection

```{r}

# calculate the RSS for the local NDVI values
ndvi_linear <- exp(ndvi_window * ndvi_coef)

# plot the habitat selection (spatial RSS)
plot(ndvi_linear, main = "RSS - linear")

# Save the plot
png(filename = paste0(plot_save_path, "/rss_linear_", which_buffalo, ".png"), 
    width = 100, height = 80, units = "mm", res = 600)
plot(ndvi_linear, main = "RSS - linear", range = c(0.5, 1.75))
dev.off()

```

## Quadratic covariate

### Model formula

$y \sim ndvi + ndvi^2$

### Fit model

```{r}

# fit an issf model
ssf_quadratic <- amt::fit_issf(case_ ~ 
                                 ndvi + 
                                 ndvi_sq + 
                                 sl_ + log_sl + cos_ta +
                              strata(step_id_),
                            data = buffalo_ssf_data)

summary(ssf_quadratic)
AIC(ssf_quadratic)

# fit an ssf gam model with parametric terms. -ends up being the same as above, 
# but allows us to use the `predict` function to get predictions on the log-scale with standard errors, 
# which is a bit more work to do with the `amt` model object.
ssf_quadratic_gam <- mgcv::gam(cbind(times, step_id_) ~ 
                                 ndvi + 
                                 ndvi_sq +
                               sl_ + log_sl + cos_ta,
                               data = buffalo_ssf_data,
                               family = cox.ph,
                               weight = case_)

summary(ssf_quadratic_gam)
AIC(ssf_quadratic_gam)

```

### Generate the response curve

```{r}

# Get predictions on log-scale with standard errors
predictions <- predict(ssf_quadratic_gam, 
                      newdata = pred_data, 
                      type = "link",  # log-scale
                      se.fit = TRUE)

# Extract log RSS and standard errors
log_rss <- predictions$fit
se_log_rss <- predictions$se.fit

# Calculate confidence intervals on log-scale
log_rss_lower <- log_rss - 1.96 * se_log_rss
log_rss_upper <- log_rss + 1.96 * se_log_rss

# Transform to RSS scale
rss <- exp(log_rss)
rss_lower <- exp(log_rss_lower)
rss_upper <- exp(log_rss_upper)

# Create your dataframe
ndvi_coef_df <- data.frame(
  ndvi_values = ndvi_values,
  log_rss = log_rss,
  log_rss_lower = log_rss_lower,
  log_rss_upper = log_rss_upper,
  rss = rss,
  rss_lower = rss_lower,
  rss_upper = rss_upper
)

# plot the response curve
ggplot(data = ndvi_coef_df) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey") +
  geom_ribbon(aes(x = ndvi_values, ymin = rss_lower, ymax = rss_upper), 
              fill = "grey70", alpha = 0.5) +
  geom_line(aes(x = ndvi_values, y = rss), 
            colour = "black", linewidth = 1) +
  geom_rug(data = buffalo_ssf_data %>% filter(case_ == T), 
           aes(x = ndvi), alpha = 0.25, sides="b") +
  scale_x_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  scale_y_continuous(limits = c(rss_min, rss_max)) +
  labs(x = "NDVI", y = "RSS") +
  theme_classic() 

ggsave(paste0(plot_save_path, "/ndvi_quadratic_response_id", which_buffalo, ".png"),
       width = 120, height = 100, units = "mm", dpi = 600)

```

### Response curve

```{r}

# extract the NDVI coefficient
ndvi_coef <- ssf_quadratic$model$coefficients[1]
ndvi_coef_sq <- ssf_quadratic$model$coefficients[2]

ndvi_se = sqrt(diag(ssf_quadratic$model$var))[1]
ndvi_sq_se = sqrt(diag(ssf_quadratic$model$var))[2]

log_rss = (ndvi_values * ndvi_coef) + (ndvi_values^2 * ndvi_coef_sq)
rss = exp(log_rss)

# calculate the RSS for each NDVI value
ndvi_coef_df <- data.frame(ndvi_values, log_rss, rss)

# plot the response curve
ggplot() +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey") +
  geom_line(data = ndvi_coef_df, aes(x = ndvi_values, y = rss), 
            colour = "black", linewidth = 1) +
  scale_x_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  labs(x = "NDVI", y = "RSS") +
  theme_classic() 

# ggsave(paste0(plot_save_path, /ndvi_quadratic_response_id", which_buffalo, ".png"),
#        width = 100, height = 100, units = "mm", dpi = 600)

```

### Plot habitat selection

```{r}

# extract the NDVI coefficient
ndvi_coef <- ssf_quadratic$model$coefficients[1]
ndvi_coef_sq <- ssf_quadratic$model$coefficients[2]

# calculate the RSS for the local NDVI values
ndvi_quadratic <- exp((ndvi_window * ndvi_coef) + 
                        (ndvi_window^2 * ndvi_coef_sq))

# plot the habitat selection (spatial RSS)
plot(ndvi_quadratic, main = "RSS - quadratic")

# Save the plot
png(filename = paste0(plot_save_path, "/rss_quadratic_", which_buffalo, ".png"), 
    width = 100, height = 80, units = "mm", res = 600)
plot(ndvi_quadratic, main = "RSS - quadratic", range = c(0.5, 1.75))
dev.off()

```

## Smooth term

Fit a GAM to the data, with a smooth term across the range of NDVI values [@Klappstein2024-ax].

### Model formula

$y \sim s(ndvi)$

### Fit model

```{r}

# fit an ssf gam model
ssf_gam <- mgcv::gam(cbind(times, step_id_) ~ 
                       s(ndvi, bs = "cr") + 
                       # s(ndvi) +
                       sl_ + log_sl + cos_ta,
                            data = buffalo_ssf_data,
                            family = cox.ph,
                            weight = case_)

summary(ssf_gam)
AIC(ssf_gam)

# coef(ssf_gam)[4]

gam_smooths <- gratia::smooth_estimates(ssf_gam)

```

Using ggplot with the `smooth_estimates` dataframe from `gratia`

```{r}

ggplot(data = gam_smooths) + 
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey") +
  geom_ribbon(aes(x = ndvi, ymin = exp(.estimate - 1.96 * .se), ymax = exp(.estimate + 1.96 * .se)), 
              fill = "grey70", alpha = 0.5) +
  geom_line(aes(x = ndvi, y = exp(.estimate)), 
            colour = "black", linewidth = 1) +
  labs(x = "NDVI", y = "RSS") +
  # scale_x_continuous(limits = c(ndvi_quantiles[1], 0.5)) +
  # scale_y_continuous(limits = c(0.65, 1.3)) +
  theme_classic() 

ggsave(paste0(plot_save_path, "/ndvi_smooth_response_id", which_buffalo, ".png"),
       width = 120, height = 100, units = "mm", dpi = 600)

```

### ALternatively, generate the response curve with the `predict` function

```{r}

# Get predictions on log-scale with standard errors
predictions <- predict(ssf_gam, 
                      newdata = pred_data, 
                      type = "link",  # log-scale
                      se.fit = TRUE)

# Extract log RSS and standard errors
log_rss <- predictions$fit
se_log_rss <- predictions$se.fit

# Calculate confidence intervals on log-scale
log_rss_lower <- log_rss - 1.96 * se_log_rss
log_rss_upper <- log_rss + 1.96 * se_log_rss

# Transform to RSS scale
rss <- exp(log_rss)
rss_lower <- exp(log_rss_lower)
rss_upper <- exp(log_rss_upper)

# Create your dataframe
ndvi_coef_df <- data.frame(
  ndvi_values = ndvi_values,
  log_rss = log_rss,
  log_rss_lower = log_rss_lower,
  log_rss_upper = log_rss_upper,
  rss = rss,
  rss_lower = rss_lower,
  rss_upper = rss_upper
)

# plot the response curve
ggplot(data = ndvi_coef_df) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey") +
  geom_ribbon(aes(x = ndvi_values, ymin = rss_lower, ymax = rss_upper), 
              fill = "grey70", alpha = 0.5) +
  geom_line(aes(x = ndvi_values, y = rss), 
            colour = "black", linewidth = 1) +
  geom_rug(data = buffalo_ssf_data %>% filter(case_ == T), 
           aes(x = ndvi), alpha = 0.25, sides="b") +
  scale_x_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  scale_y_continuous(limits = c(rss_min, rss_max)) +
  labs(x = "NDVI", y = "RSS") +
  theme_classic() 

ggsave(paste0(plot_save_path, "/ndvi_smooth_response_id", which_buffalo, ".png"),
       width = 120, height = 100, units = "mm", dpi = 600)

```

### Plot habitat selection

```{r}

predicted_log_rss <- predict(ssf_gam, 
                         newdata = data.frame(ndvi = values(ndvi_window), 
                                                sl_ = 0, 
                                                log_sl = 0, 
                                                cos_ta = 0),, 
                         type = "link")

ndvi_gam <- ndvi_window
values(ndvi_gam) <- exp(as.vector(predicted_log_rss))

# Plot the predictions
plot(ndvi_gam, main = "RSS - smooth term")

# Save the plot
png(filename = paste0(plot_save_path, "/rss_smooth_", which_buffalo, ".png"), 
    width = 100, height = 80, units = "mm", res = 600)
plot(ndvi_gam, main = "RSS - smooth", range = c(0.5, 1.75))
dev.off()

```

# Temporal dynamics

## Static model for comparison

```{r}

ssf_static <- amt::fit_issf(case_ ~ 
                                 ndvi +sl_ + log_sl + cos_ta +
                                 strata(step_id_),
                               data = buffalo_ssf_data)

summary(ssf_static)
AIC(ssf_static)

```

```{r}

ndvi_coef <- ssf_static$model$coefficients["ndvi"]
ndvi_se <- sqrt(diag(ssf_static$model$var))[1]  # SE of coefficient

ndvi_static_coef_df <- data.frame(hour = seq(0, 23, length.out = 231),
                                    ndvi_coef = ndvi_coef,
                                    ndvi_se = ndvi_se,
                                    ndvi_lo = ndvi_coef - 1.96 * ndvi_se,
                                    ndvi_hi = ndvi_coef + 1.96 * ndvi_se)

# Plot with ribbon
ggplot(ndvi_static_coef_df, aes(x = hour)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "black") +
  geom_vline(xintercept = 9, linetype = "dashed", colour = "orange") +
  geom_vline(xintercept = 17, linetype = "dashed", colour = "skyblue") +
  geom_ribbon(aes(ymin = ndvi_lo, ymax = ndvi_hi), alpha = 0.2) +
  geom_line(aes(y = ndvi_coef)) +
  labs(title = "ndvi", x = "Hour", y = expression(beta[NDVI]), fill = NULL) +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  scale_y_continuous(limits = c(-5,3.5), breaks = seq(-5, 4, by = 1)) +
  theme_classic()

ggsave(paste0(plot_save_path, "/ndvi_static_coef_response_id", which_buffalo, ".png"),
       width = 150, height = 80, units = "mm", dpi = 600)

```

### Pull out a response curve from one of the hours

```{r}

sample_hours <- c(9, 17)

ndvi_coef_df <- purrr::map_dfr(sample_hours, function(sample_hour) {
  
  # extract the NDVI coefficient and its standard error
  ndvi_coef <- ndvi_static_coef_df %>% filter(hour == sample_hour) %>% pull(ndvi_coef)
  ndvi_coef_se <- ndvi_static_coef_df %>% filter(hour == sample_hour) %>% pull(ndvi_se)
  
  # calculate the RSS for each NDVI value
  data.frame(
    ndvi_values = ndvi_values,
    log_rss = ndvi_values * ndvi_coef,
    rss = exp(ndvi_values * ndvi_coef),
    se_prediction = abs(ndvi_values) * ndvi_coef_se
  ) %>% mutate(hour = as.factor(sample_hour), .before = ndvi_values)
  
})

# plot the response curve
ggplot(data = ndvi_coef_df) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey") +
  geom_ribbon(aes(x = ndvi_values, 
                  ymin = exp(log_rss - 1.96 * se_prediction), 
                  ymax = exp(log_rss + 1.96 * se_prediction),
                  fill = hour), 
              alpha = 0.5) +
  geom_line(aes(x = ndvi_values, y = rss, colour = hour), 
            linewidth = 1) +
  geom_rug(data = buffalo_ssf_data %>% filter(case_ == T), 
           aes(x = ndvi), alpha = 0.25, sides="b") +
  scale_colour_manual(values = c("orange", "skyblue")) +
  scale_fill_manual(values = c("orange", "skyblue")) +
  scale_x_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  scale_y_continuous(limits = c(0, 5.5)) +
  labs(x = "NDVI", y = "RSS") +
  theme_classic() +
  theme(legend.position = "none")

ggsave(paste0(plot_save_path, "/ndvi_static_response_sample_hour_id", which_buffalo, ".png"),
       width = 70, height = 80, units = "mm", dpi = 600)

```

## Fit harmonic model with quadratic terms

```{r}

ssf_static_quadratic <- amt::fit_issf(case_ ~ 
                                 # linear term
                                 ndvi + 
                                 # quadratic terms
                                 ndvi_sq +
                                 sl_ + log_sl + cos_ta +
                                 strata(step_id_),
                               data = buffalo_ssf_data)

summary(ssf_static_quadratic)
AIC(ssf_static_quadratic)

```

### Create table of NDVI and hour values

```{r}

ndvi_hour_values <- expand.grid(ndvi = ndvi_values, 
                                hour = seq(0, 23, length.out = length(ndvi_values)))

ndvi_lin_coef <- ssf_static_quadratic$model$coefficients["ndvi"]
ndvi_quad_coef <- ssf_static_quadratic$model$coefficients["ndvi_sq"]

ndvi_hour_values <- ndvi_hour_values %>% mutate(
  ndvi_lin_coef = ndvi_lin_coef,
  ndvi_quad_coef = ndvi_quad_coef,
  ssf_quadratic_log_rss = (ndvi * ndvi_lin_coef) + (ndvi^2 * ndvi_quad_coef)
)


max_abs_colour <- max(abs(range(ndvi_hour_values$ssf_quadratic_log_rss)))

# plot the selection surface
ggplot() +
  geom_raster(data = ndvi_hour_values,
       aes(x = hour, y = ndvi, fill = ssf_quadratic_log_rss)) +
  labs(title = "(ndvi + ndvi^2)", x = "Hour", y = "NDVI", fill = "log-RSS") +
  geom_contour(data = ndvi_hour_values,
       aes(x = hour, y = ndvi, z = ssf_quadratic_log_rss), 
       colour = "black") +
  geom_contour(data = ndvi_hour_values,
       aes(x = hour, y = ndvi, z = ssf_quadratic_log_rss),
       breaks = 0, colour = "black", linetype = "solid", linewidth = 0.75) +
  geom_vline(xintercept = 9, linetype = "dashed", colour = "orange") +
  geom_vline(xintercept = 17, linetype = "dashed", colour = "skyblue") +
  scale_fill_distiller(palette = "RdBu", type = "div", direction = 1,
                       limits = c(-max_abs_colour, max_abs_colour)) +
  coord_cartesian(expand = FALSE) +
  geom_rug(data = buffalo_ssf_data %>% filter(case_ == T),
           aes(y = ndvi), alpha = 0.25, length = unit(0.015, "npc"), sides="l") +
  scale_y_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  theme_bw() 

ggsave(paste0(plot_save_path, "/ndvi_static_quadratic_response_id", which_buffalo, ".png"),
       width = 150, height = 80, units = "mm", dpi = 600)

```

### Easier to get the curves with GAM predict function

### Fit model

```{r}

# fit an ssf gam model with parametric terms. -ends up being the same as above, 
# but allows us to use the `predict` function to get predictions on the log-scale with standard errors, 
# which is a bit more work to do with the `amt` model object.
ssf_quadratic_gam <- mgcv::gam(cbind(times, step_id_) ~ 
                                 ndvi + 
                                 ndvi_sq +
                               sl_ + log_sl + cos_ta,
                               data = buffalo_ssf_data,
                               family = cox.ph,
                               weight = case_)

summary(ssf_quadratic_gam)
AIC(ssf_quadratic_gam)

```

### Generate the response curve

```{r}

# Get predictions on log-scale with standard errors
predictions <- predict(ssf_quadratic_gam, 
                      newdata = pred_data, 
                      type = "link",  # log-scale
                      se.fit = TRUE)

# Extract log RSS and standard errors
log_rss <- predictions$fit
se_log_rss <- predictions$se.fit

# Calculate confidence intervals on log-scale
log_rss_lower <- log_rss - 1.96 * se_log_rss
log_rss_upper <- log_rss + 1.96 * se_log_rss

# Transform to RSS scale
rss <- exp(log_rss)
rss_lower <- exp(log_rss_lower)
rss_upper <- exp(log_rss_upper)

# Create your dataframe
ndvi_coef_df <- data.frame(
  ndvi_values = ndvi_values,
  log_rss = log_rss,
  log_rss_lower = log_rss_lower,
  log_rss_upper = log_rss_upper,
  rss = rss,
  rss_lower = rss_lower,
  rss_upper = rss_upper
)

ndvi_coef_df <- rbind(data.frame(hour = 9, ndvi_coef_df), data.frame(hour = 17, ndvi_coef_df))

# plot the response curve
ggplot(data = ndvi_coef_df) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey") +
  geom_ribbon(aes(x = ndvi_values, ymin = rss_lower, ymax = rss_upper, fill = as.factor(hour)), 
              alpha = 0.5) +
  geom_line(aes(x = ndvi_values, y = rss, colour = as.factor(hour)), 
            linewidth = 1) +
  geom_rug(data = buffalo_ssf_data %>% filter(case_ == T), 
           aes(x = ndvi), alpha = 0.25, sides="b") +
  scale_colour_manual(values = c("orange", "skyblue")) +
  scale_fill_manual(values = c("orange", "skyblue")) +
  scale_x_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  scale_y_continuous(limits = c(0, 5.5)) +
  labs(x = "NDVI", y = "RSS") +
  theme_classic()  +
  theme(legend.position = "none")

ggsave(paste0(plot_save_path, "/ndvi_static_quadratic_response_sample_hour_id", which_buffalo, ".png"),
       width = 70, height = 80, units = "mm", dpi = 600)

```


## Harmonic regression

```{r}

buffalo_ssf_data <- buffalo_ssf_data %>% mutate(
  
  # create hour term
  hour = lubridate::hour(t1_),
  
  # harmonic terms
  sin_1_hour = sin(2 * pi * hour / 24),
  cos_1_hour = cos(2 * pi * hour / 24),
  sin_2_hour = sin(4 * pi * hour / 24),
  cos_2_hour = cos(4 * pi * hour / 24),
  
)

```

### Fit model

We'll use the `fit_issf` function from `amt`, with manually specified harmonic terms.

```{r}

ssf_harmonics <- amt::fit_issf(case_ ~ 
                                 ndvi +
                                 ndvi:sin_1_hour + ndvi:cos_1_hour +
                                 ndvi:sin_2_hour + ndvi:cos_2_hour +
                                 sl_ + log_sl + cos_ta +
                                 strata(step_id_),
                               data = buffalo_ssf_data)

summary(ssf_harmonics)
AIC(ssf_harmonics)

```

### Pull out the coefficent across time

```{r}

ndvi_harmonic_coef_reconstruction <- function(ssf_harmonic_model, hour) {
  
  ndvi_coef <- ssf_harmonic_model$model$coefficients["ndvi"]
  ndvi_sin_1_coef <- ssf_harmonic_model$model$coefficients["ndvi:sin_1_hour"]
  ndvi_cos_1_coef <- ssf_harmonic_model$model$coefficients["ndvi:cos_1_hour"]
  ndvi_sin_2_coef <- ssf_harmonic_model$model$coefficients["ndvi:sin_2_hour"]
  ndvi_cos_2_coef <- ssf_harmonic_model$model$coefficients["ndvi:cos_2_hour"]
  
  sin_1_hour <- sin(2 * pi * hour / 24)
  cos_1_hour <- cos(2 * pi * hour / 24)
  sin_2_hour <- sin(4 * pi * hour / 24)
  cos_2_hour <- cos(4 * pi * hour / 24)
  
  coef <- as.numeric(ndvi_coef + 
                       sin_1_hour * ndvi_sin_1_coef + 
                       cos_1_hour * ndvi_cos_1_coef + 
                       sin_2_hour * ndvi_sin_2_coef + 
                       cos_2_hour * ndvi_cos_2_coef)
  
  return(coef)
  
}

ndvi_temporal_coef_df <- data.frame(hour = seq(0, 23, length.out = 100))

ndvi_temporal_coef_df$ndvi_coef <- sapply(ndvi_temporal_coef_df$hour, 
                                            ndvi_harmonic_coef_reconstruction, 
                                            ssf_harmonic_model = ssf_harmonics)

# plot the response curve
ggplot(data = ndvi_temporal_coef_df) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey") +
  geom_line(aes(x = hour, y = ndvi_coef), 
            colour = "black") +
  labs(x = "Hour", y = "NDVI coefficient") +
  theme_classic()

```

### Calculate confidence intervals across time and plot

```{r}

sample_hour <- 9 # hour to plot with a vertical dashed line

# Coefficient names we need
coef_names <- c("ndvi",
                "ndvi:sin_1_hour","ndvi:cos_1_hour",
                "ndvi:sin_2_hour","ndvi:cos_2_hour")

# coefficients and V from the underlying clogit model
beta <- coef(ssf_harmonics$model)[coef_names]
V    <- vcov(ssf_harmonics$model)[coef_names, coef_names, drop = FALSE]

# Build design for each hour
ndvi_temporal_ci <- tibble(hour = seq(0, 23, length.out = 231)) %>%
  mutate(
    sin_1 = sin(2 * pi * hour / 24),
    cos_1 = cos(2 * pi * hour / 24),
    sin_2 = sin(4 * pi * hour / 24),
    cos_2 = cos(4 * pi * hour / 24)
  ) %>%
  rowwise() %>%
  mutate(
    x = list(c(1, sin_1, cos_1, sin_2, cos_2)),
    est = as.numeric(matrix(unlist(x), nrow = 1) %*% beta),
    se  = sqrt(as.numeric(matrix(unlist(x), nrow = 1) %*% V %*% matrix(unlist(x), ncol = 1))),
    lo  = est - 1.96 * se,
    hi  = est + 1.96 * se
  ) %>%
  ungroup() %>%
  select(hour, est, se, lo, hi)

# Plot with ribbon
ggplot(ndvi_temporal_ci, aes(x = hour)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "black") +
  geom_vline(xintercept = 9, linetype = "dashed", colour = "orange") +
  geom_vline(xintercept = 17, linetype = "dashed", colour = "skyblue") +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2) +
  geom_line(aes(y = est)) +
  labs(title = "harmonics(ndvi)", x = "Hour", y = expression(beta[NDVI]), fill = NULL) +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  scale_y_continuous(limits = c(-5,3.5), breaks = seq(-5, 4, by = 1)) +
  theme_classic()

ggsave(paste0(plot_save_path, "/ndvi_harmonic_coef_response_id", which_buffalo, ".png"),
       width = 150, height = 80, units = "mm", dpi = 600)

```

### Pull out a response curve from one of the hours

```{r}

sample_hours <- c(9, 17)

ndvi_coef_df <- purrr::map_dfr(sample_hours, function(sample_hour) {
  
  # extract the NDVI coefficient and its standard error
  ndvi_coef <- ndvi_temporal_ci %>% filter(hour == sample_hour) %>% pull(est)
  ndvi_coef_se <- ndvi_temporal_ci %>% filter(hour == sample_hour) %>% pull(se)
  
  # calculate the RSS for each NDVI value
  data.frame(
    ndvi_values = ndvi_values,
    log_rss = ndvi_values * ndvi_coef,
    rss = exp(ndvi_values * ndvi_coef),
    se_prediction = abs(ndvi_values) * ndvi_coef_se
  ) %>% mutate(hour = as.factor(sample_hour), .before = ndvi_values)
})

# plot the response curve
ggplot(data = ndvi_coef_df) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey") +
  geom_ribbon(aes(x = ndvi_values, 
                  ymin = exp(log_rss - 1.96 * se_prediction), 
                  ymax = exp(log_rss + 1.96 * se_prediction),
                  fill = hour), 
              alpha = 0.5) +
  geom_line(aes(x = ndvi_values, y = rss, colour = hour), 
            linewidth = 1) +
  geom_rug(data = buffalo_ssf_data %>% filter(case_ == T), 
           aes(x = ndvi), alpha = 0.25, sides="b") +
  scale_colour_manual(values = c("orange", "skyblue")) +
  scale_fill_manual(values = c("orange", "skyblue")) +
  scale_x_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  scale_y_continuous(limits = c(0, 5.5)) +
  labs(x = "NDVI", y = "RSS") +
  theme_classic() +
  theme(legend.position = "none")

ggsave(paste0(plot_save_path, "/ndvi_linear_response_sample_hour_id", which_buffalo, ".png"),
       width = 70, height = 80, units = "mm", dpi = 600)

```

### Manually create the selection surface

```{r}

ndvi_harmonic_reconstruction <- function(ssf_harmonic_model, ndvi_value, hour) {
  
  ndvi_coef <- ssf_harmonic_model$model$coefficients["ndvi"]
  ndvi_sin_1_coef <- ssf_harmonic_model$model$coefficients["ndvi:sin_1_hour"]
  ndvi_cos_1_coef <- ssf_harmonic_model$model$coefficients["ndvi:cos_1_hour"]
  ndvi_sin_2_coef <- ssf_harmonic_model$model$coefficients["ndvi:sin_2_hour"]
  ndvi_cos_2_coef <- ssf_harmonic_model$model$coefficients["ndvi:cos_2_hour"]
  
  sin_1_hour <- sin(2 * pi * hour / 24)
  cos_1_hour <- cos(2 * pi * hour / 24)
  sin_2_hour <- sin(4 * pi * hour / 24)
  cos_2_hour <- cos(4 * pi * hour / 24)
  
  log_rss <- as.numeric(ndvi_value * ndvi_coef + 
                          ndvi_value * sin_1_hour * ndvi_sin_1_coef + 
                          ndvi_value * cos_1_hour * ndvi_cos_1_coef + 
                          ndvi_value * sin_2_hour * ndvi_sin_2_coef + 
                          ndvi_value * cos_2_hour * ndvi_cos_2_coef)
  
  return(log_rss)
  
}

# test function
ndvi_harmonic_reconstruction(ssf_harmonics, 
                             ndvi_values[1], 
                             hour = 12)

```

### Create table of NDVI and hour values

```{r}

ndvi_hour_values <- expand.grid(ndvi = ndvi_values, 
                                hour = seq(0, 23, length.out = length(ndvi_values)))

ndvi_hour_values$ssf_linear_log_rss <- mapply(ndvi_harmonic_reconstruction, 
                                        ssf_harmonic_model = list(ssf_harmonics), 
                                        ndvi_value = ndvi_hour_values$ndvi, 
                                        hour = ndvi_hour_values$hour)

# plot the response curve
ggplot(data = ndvi_hour_values) +
  geom_point(aes(x = hour, y = ndvi, colour = ssf_linear_log_rss)) +
  labs(x = "Hour", y = "NDVI", colour = "log RSS") +
  scale_colour_viridis_c() +
  theme_classic() 

max_abs_colour <- max(abs(range(ndvi_hour_values$ssf_linear_log_rss)))

# plot the selection surface
ggplot() +
  geom_raster(data = ndvi_hour_values,
       aes(x = hour, y = ndvi, fill = ssf_linear_log_rss)) +
  labs(title = "harmonics(ndvi)", x = "Hour", y = "NDVI", fill = "log-RSS") +
  geom_contour(data = ndvi_hour_values,
       aes(x = hour, y = ndvi, z = ssf_linear_log_rss), 
       colour = "black") +
  geom_contour(data = ndvi_hour_values,
       aes(x = hour, y = ndvi, z = ssf_linear_log_rss),
       breaks = 0, colour = "black", linetype = "solid", linewidth = 1) +
  scale_fill_distiller(palette = "RdBu", type = "div", direction = 1,
    limits = c(-max_abs_colour, max_abs_colour)) +   # symmetric around 0) 
  coord_cartesian(expand = FALSE) +
  geom_rug(data = buffalo_ssf_data %>% filter(case_ == T),
           aes(y = ndvi), alpha = 0.1, length = unit(0.015, "npc"), sides="l") +
  scale_y_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  theme_bw()

ggsave(paste0(plot_save_path, "/ndvi_harmonic_response_id", which_buffalo, ".png"),
       width = 150, height = 80, units = "mm", dpi = 600)

```

## Fit harmonic model with quadratic terms

```{r}

ssf_harmonics_quadratic <- amt::fit_issf(case_ ~ 
                                 
                                 # linear terms
                                 ndvi + 
                                 ndvi:sin_1_hour + ndvi:cos_1_hour +
                                 ndvi:sin_2_hour + ndvi:cos_2_hour +
                                 
                                 # quadratic terms
                                 ndvi_sq +
                                 ndvi_sq:sin_1_hour + ndvi_sq:cos_1_hour +
                                 ndvi_sq:sin_2_hour + ndvi_sq:cos_2_hour +
                                 
                                 sl_ + log_sl + cos_ta +
                                   
                                 strata(step_id_),
                               data = buffalo_ssf_data)

summary(ssf_harmonics_quadratic)
AIC(ssf_harmonics_quadratic)

```

### Manually create the selection surface

```{r}

ndvi_harmonic_reconstruction <- function(ssf_harmonic_model, ndvi_value, hour) {
  
  # linear terms
  ndvi_coef <- ssf_harmonic_model$model$coefficients["ndvi"]
  ndvi_sin_1_coef <- ssf_harmonic_model$model$coefficients["ndvi:sin_1_hour"]
  ndvi_cos_1_coef <- ssf_harmonic_model$model$coefficients["ndvi:cos_1_hour"]
  ndvi_sin_2_coef <- ssf_harmonic_model$model$coefficients["ndvi:sin_2_hour"]
  ndvi_cos_2_coef <- ssf_harmonic_model$model$coefficients["ndvi:cos_2_hour"]
  
  # quadratic terms
  ndvi_sq_coef <- ssf_harmonic_model$model$coefficients["ndvi_sq"]
  ndvi_sq_sin_1_coef <- ssf_harmonic_model$model$coefficients["ndvi_sq:sin_1_hour"]
  ndvi_sq_cos_1_coef <- ssf_harmonic_model$model$coefficients["ndvi_sq:cos_1_hour"]
  ndvi_sq_sin_2_coef <- ssf_harmonic_model$model$coefficients["ndvi_sq:sin_2_hour"]
  ndvi_sq_cos_2_coef <- ssf_harmonic_model$model$coefficients["ndvi_sq:cos_2_hour"]
  
  sin_1_hour <- sin(2 * pi * hour / 24)
  cos_1_hour <- cos(2 * pi * hour / 24)
  sin_2_hour <- sin(4 * pi * hour / 24)
  cos_2_hour <- cos(4 * pi * hour / 24)
  
  linear_part <- ndvi_value * ndvi_coef + 
    ndvi_value * sin_1_hour * ndvi_sin_1_coef + 
    ndvi_value * cos_1_hour * ndvi_cos_1_coef + 
    ndvi_value * sin_2_hour * ndvi_sin_2_coef + 
    ndvi_value * cos_2_hour * ndvi_cos_2_coef
  
  ndvi_sq_value <- ndvi_value^2
  
  quadratic_part <- ndvi_sq_value * ndvi_sq_coef + 
    ndvi_sq_value * sin_1_hour * ndvi_sq_sin_1_coef + 
    ndvi_sq_value * cos_1_hour * ndvi_sq_cos_1_coef + 
    ndvi_sq_value * sin_2_hour * ndvi_sq_sin_2_coef + 
    ndvi_sq_value * cos_2_hour * ndvi_sq_cos_2_coef
  
  return(as.numeric(linear_part + quadratic_part))
  
}

# test function
ndvi_harmonic_reconstruction(ssf_harmonics_quadratic, ndvi_values[1], hour = 12)

```

### Create table of NDVI and hour values

```{r}

ndvi_hour_values <- expand.grid(ndvi = ndvi_values, 
                                hour = seq(0, 23, length.out = length(ndvi_values)))

ndvi_hour_values$ssf_quadratic_log_rss <- mapply(ndvi_harmonic_reconstruction, 
                                                 ssf_harmonic_model = list(ssf_harmonics_quadratic), 
                                                 ndvi_value = ndvi_hour_values$ndvi, 
                                                 hour = ndvi_hour_values$hour)

# plot the response curve
ggplot(data = ndvi_hour_values) +
  geom_point(aes(x = hour, y = ndvi, colour = ssf_quadratic_log_rss)) +
  labs(x = "Hour", y = "NDVI", colour = "log RSS") +
  scale_colour_viridis_c() +
  scale_y_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  theme_classic() 

max_abs_colour <- max(abs(range(ndvi_hour_values$ssf_quadratic_log_rss)))

# plot the selection surface
ggplot() +
  geom_raster(data = ndvi_hour_values,
       aes(x = hour, y = ndvi, fill = ssf_quadratic_log_rss)) +
  labs(title = "harmonics(ndvi + ndvi^2)", x = "Hour", y = "NDVI", fill = "log-RSS") +
  geom_contour(data = ndvi_hour_values,
       aes(x = hour, y = ndvi, z = ssf_quadratic_log_rss), 
       colour = "black") +
  geom_contour(data = ndvi_hour_values,
       aes(x = hour, y = ndvi, z = ssf_quadratic_log_rss),
       breaks = 0, colour = "black", linetype = "solid", linewidth = 1) +
  geom_vline(xintercept = 9, linetype = "dashed", colour = "orange") +
  geom_vline(xintercept = 17, linetype = "dashed", colour = "skyblue") +
  scale_fill_distiller(palette = "RdBu", type = "div", direction = 1,
                       limits = c(-max_abs_colour, max_abs_colour)) +
  coord_cartesian(expand = FALSE) +
  geom_rug(data = buffalo_ssf_data %>% filter(case_ == T),
           aes(y = ndvi), alpha = 0.25, length = unit(0.015, "npc"), sides="l") +
  scale_y_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  theme_bw() 

ggsave(paste0(plot_save_path, "/ndvi_harmonic_quadratic_response_id", which_buffalo, ".png"),
       width = 150, height = 80, units = "mm", dpi = 600)

```

### Pull out a response curve from one of the hours

```{r}

sample_hours <- c(9, 17)

curve_df <- purrr::map_dfr(sample_hours, function(sample_hour) {
  
  # Harmonics at sample hour
  s1 <- sin(2 * pi * sample_hour / 24)
  c1 <- cos(2 * pi * sample_hour / 24)
  s2 <- sin(4 * pi * sample_hour / 24)
  c2 <- cos(4 * pi * sample_hour / 24)
  
  mod   <- ssf_harmonics_quadratic$model
  betas <- coef(mod)
  V     <- vcov(mod)
  cn    <- names(betas)
  
  # For each NDVI value, build the contrast vector g such that g'β = log-RSS
  purrr::map_dfr(ndvi_values, function(x) {
    
    g <- setNames(numeric(length(betas)), cn)
    
    # linear term: x * (β_ndvi + harmonics)
    g["ndvi"]            <- x
    g["ndvi:sin_1_hour"] <- x * s1
    g["ndvi:cos_1_hour"] <- x * c1
    g["ndvi:sin_2_hour"] <- x * s2
    g["ndvi:cos_2_hour"] <- x * c2
    
    # quadratic term: x² * (β_ndvi_sq + harmonics)
    g["ndvi_sq"]             <- x^2
    g["ndvi_sq:sin_1_hour"]  <- x^2 * s1
    g["ndvi_sq:cos_1_hour"]  <- x^2 * c1
    g["ndvi_sq:sin_2_hour"]  <- x^2 * s2
    g["ndvi_sq:cos_2_hour"]  <- x^2 * c2
    
    log_rss <- as.numeric(g %*% betas)
    se_log  <- sqrt(as.numeric(t(g) %*% V %*% g))
    
    data.frame(
      hour    = sample_hour,
      ndvi    = x,
      log_rss = log_rss,
      rss     = exp(log_rss),
      # delta method on the RSS scale: SE_rss ≈ exp(log_rss) * SE_log
      rss_lwr = exp(log_rss - 1.96 * se_log),
      rss_upr = exp(log_rss + 1.96 * se_log)
    )
  })
})


# plot the response curve
ggplot(curve_df, aes(x = ndvi)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey") +
  geom_ribbon(aes(ymin = rss_lwr, ymax = rss_upr, fill = as.factor(hour)), 
              alpha = 0.5) +
  geom_line(aes(y = rss, colour = as.factor(hour)), linewidth = 1) +
  geom_rug(data = buffalo_ssf_data %>% filter(case_ == T), 
           aes(x = ndvi), alpha = 0.25, sides="b") +
  scale_colour_manual(values = c("orange", "skyblue")) +
  scale_fill_manual(values = c("orange", "skyblue")) +
  scale_x_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) + 
  scale_y_continuous(limits = c(0, 5.5)) +
  labs(x = "NDVI", y = "RSS") +
  theme_classic() +
  theme(legend.position = "none")

ggsave(paste0(plot_save_path, "/ndvi_quadratic_response_sample_hour_id", which_buffalo, ".png"),
       width = 70, height = 80, units = "mm", dpi = 600)

```

## Temporal dynamics with smooth terms

```{r}

# fit an ssf gam model
# We set the knots explicitly at 0 and 24. Without this, mgcv places the cyclic
# endpoints at the range of the data (0 and 23), which wraps hour 23 onto hour 0
# and squeezes an hour out of the daily cycle.
ssf_gam_temporal <- mgcv::gam(cbind(times, step_id_) ~
                       s(hour, by = ndvi, bs = "cc") + #, k = 5
                       sl_ + log_sl + cos_ta,
                     knots = list(hour = c(0, 24)),
                     data = buffalo_ssf_data,
                     family = cox.ph,
                     weight = case_)

summary(ssf_gam_temporal)
AIC(ssf_gam_temporal)

gam_smooths <- smooth_estimates(ssf_gam_temporal)

```


```{r}

gratia::draw(ssf_gam_temporal, rug = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 9, linetype = "dashed", colour = "orange") +
  geom_vline(xintercept = 17, linetype = "dashed", colour = "skyblue") +
  labs() +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  scale_y_continuous(limits = c(-5,3.5), breaks = seq(-5, 4, by = 1)) +
  labs(title = "s(hour, by = ndvi)",
       subtitle = NULL,
       caption = NULL,
       x = "Hour", 
       y = expression(beta[NDVI]), fill = NULL) +
  theme_classic()

ggsave(paste0(plot_save_path, "/ndvi_gam_coef_response_id", which_buffalo, ".png"),
       width = 150, height = 80, units = "mm", dpi = 600)

```

### Pull out a response curve from one of the hours

```{r}

pred_data_hour9 <- pred_data
pred_data_hour9$hour <- 9

pred_data_hour17 <- pred_data
pred_data_hour17$hour <- 17

pred_data_hour9_17 <- bind_rows(pred_data_hour9, pred_data_hour17)

# Get predictions on log-scale with standard errors
predictions <- predict(ssf_gam_temporal, 
                      newdata = pred_data_hour9_17, 
                      type = "link",  # log-scale
                      se.fit = TRUE)

pred_data_hour9_17 <- pred_data_hour9_17 %>% 
  mutate(
    # Extract log RSS and standard errors
    log_rss = predictions$fit,
    se_log_rss = predictions$se.fit,
    # Calculate confidence intervals on log-scale
    log_rss_lower = log_rss - 1.96 * se_log_rss,
    log_rss_upper = log_rss + 1.96 * se_log_rss,
    # Transform to RSS scale
    rss = exp(log_rss),
    rss_lower = exp(log_rss_lower),
    rss_upper = exp(log_rss_upper)
  )

# plot the response curve
ggplot(data = pred_data_hour9_17) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey") +
  geom_ribbon(aes(x = ndvi, 
                  ymin = rss_lower, ymax = rss_upper,
                  fill = as.factor(hour)), 
              alpha = 0.5) +
  geom_line(aes(x = ndvi, y = rss, colour = as.factor(hour)), 
            linewidth = 1) +
  geom_rug(data = buffalo_ssf_data %>% filter(case_ == T), 
           aes(x = ndvi), alpha = 0.25, sides="b") +
  scale_colour_manual(values = c("orange", "skyblue")) +
  scale_fill_manual(values = c("orange", "skyblue")) +
  scale_x_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  scale_y_continuous(limits = c(0, 5.5)) +
  labs(x = "NDVI", y = "RSS") +
  theme_classic() +
  theme(legend.position = "none")

ggsave(paste0(plot_save_path, "/ndvi_gam_response_sample_hour_id", which_buffalo, ".png"),
       width = 70, height = 80, units = "mm", dpi = 600)

```

### GAM selection surface

```{r}

ndvi_hour_values <- expand.grid(ndvi = ndvi_values, 
                                hour = seq(0, 23, length.out = length(ndvi_values)))

ndvi_hour_values <- ndvi_hour_values %>% mutate(
  sl_ = 0,
  log_sl = 0,
  cos_ta = 0
)

ndvi_hour_values

# Get predictions on log-scale with standard errors
predictions <- predict(ssf_gam_temporal, 
                      newdata = ndvi_hour_values, 
                      type = "link",  # log-scale
                      se.fit = TRUE)

# Extract log RSS and standard errors
gam_log_rss <- predictions$fit
gam_se_log_rss <- predictions$se.fit

# Calculate confidence intervals on log-scale
gam_log_rss_lower <- gam_log_rss - 1.96 * gam_se_log_rss
gam_log_rss_upper <- gam_log_rss + 1.96 * gam_se_log_rss

# Transform to RSS scale
rss <- exp(gam_log_rss)
rss_lower <- exp(gam_log_rss_lower)
rss_upper <- exp(gam_log_rss_upper)

ndvi_hour_values <- ndvi_hour_values %>% mutate(
  gam_log_rss = gam_log_rss,
  gam_rss = exp(gam_log_rss),
  gam_log_rss_lower = as.numeric(gam_log_rss_lower),
  gam_log_rss_upper = as.numeric(gam_log_rss_upper)
)

max_abs_colour <- max(abs(range(ndvi_hour_values$gam_log_rss)))

# plot the selection surface
ggplot() +
  geom_raster(data = ndvi_hour_values,
       aes(x = hour, y = ndvi, fill = gam_log_rss)) +
  labs(title = "s(hour, by = ndvi)", x = "Hour", y = "NDVI", fill = "log-RSS") +
  geom_contour(data = ndvi_hour_values,
       aes(x = hour, y = ndvi, z = gam_log_rss), colour = "black") +
  scale_fill_distiller(palette = "RdBu", type = "div", direction = 1,
                       limits = c(-max_abs_colour, max_abs_colour)) +
  coord_cartesian(expand = FALSE) +
  geom_rug(data = buffalo_ssf_data %>% filter(case_ == T),
           aes(y = ndvi), alpha = 0.25, length = unit(0.015, "npc"), sides="l") +
  scale_y_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  theme_bw() 

ggsave(paste0(plot_save_path, "/ndvi_linear_gam_response_id", which_buffalo, ".png"),
       width = 150, height = 80, units = "mm", dpi = 600)

```

## Fit 2D GAM smooth

```{r}

# fit an ssf gam model
# As above, set the cyclic knots for the hour margin explicitly at 0 and 24
ssf_gam2d_temporal2 <- mgcv::gam(cbind(times, step_id_) ~
                                  te(hour, ndvi, bs = c("cc", "cr")) + #, k = c(5, 5)
                       sl_ + log_sl + cos_ta,
                     knots = list(hour = c(0, 24)),
                     data = buffalo_ssf_data,
                     family = cox.ph,
                     weight = case_)

summary(ssf_gam2d_temporal2)
AIC(ssf_gam2d_temporal2)

```

### Plot the smooth selection surface

```{r}

gratia::draw(ssf_gam2d_temporal2, rug = FALSE)

gam_smooths <- smooth_estimates(ssf_gam2d_temporal2)
filtered_gam_smooths <- gam_smooths %>% filter(ndvi > ndvi_quantiles[1] & ndvi < ndvi_quantiles[2])

gratia::draw(filtered_gam_smooths) +
  coord_cartesian(expand = FALSE) +
  scale_y_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  labs(x = "Hour", y = "NDVI") +
  theme_bw()


ndvi_hour_values <- expand.grid(ndvi = ndvi_values, 
                                hour = seq(0, 23, length.out = length(ndvi_values)))

ndvi_hour_values <- ndvi_hour_values %>% mutate(
  sl_ = 0,
  log_sl = 0,
  cos_ta = 0
)

predictions <- predict(ssf_gam2d_temporal2, 
                      newdata = ndvi_hour_values, 
                      type = "link",  # log-scale
                      se.fit = TRUE)

# Extract log RSS and standard errors
gam_log_rss <- predictions$fit
gam_se_log_rss <- predictions$se.fit

# Calculate confidence intervals on log-scale
gam_log_rss_lower <- gam_log_rss - 1.96 * gam_se_log_rss
gam_log_rss_upper <- gam_log_rss + 1.96 * gam_se_log_rss

ndvi_hour_values <- ndvi_hour_values %>% mutate(
  gam_log_rss = gam_log_rss,
  gam_rss = exp(gam_log_rss),
  gam_log_rss_lower = as.numeric(gam_log_rss_lower),
  gam_log_rss_upper = as.numeric(gam_log_rss_upper)
)

# plot the selection surface
ggplot(data = ndvi_hour_values) +
  geom_point(aes(x = hour, y = ndvi, colour = gam_log_rss)) +
  labs(x = "Hour", y = "NDVI", colour = "log RSS") +
  scale_colour_viridis_c() +
  scale_y_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  theme_classic() 


gratia::draw(filtered_gam_smooths) +
  coord_cartesian(expand = FALSE) +
  scale_y_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  labs(x = "Hour", y = "NDVI") +
  theme_bw()


max_abs_colour <- max(abs(range(ndvi_hour_values$gam_log_rss)))

# plot the selection surface
ggplot() +
  geom_raster(data = ndvi_hour_values,
       aes(x = hour, y = ndvi, fill = gam_log_rss)) +
  labs(title = "te(hour, ndvi)", x = "Hour", y = "NDVI", fill = "log-RSS") +
  geom_contour(data = ndvi_hour_values,
       aes(x = hour, y = ndvi, z = gam_log_rss), colour = "black") +
  geom_vline(xintercept = 9, linetype = "dashed", colour = "orange") +
  geom_vline(xintercept = 17, linetype = "dashed", colour = "skyblue") +
  scale_fill_distiller(palette = "RdBu", type = "div", direction = 1,
                       limits = c(-max_abs_colour, max_abs_colour)) +
  coord_cartesian(expand = FALSE) +
  geom_rug(data = buffalo_ssf_data %>% filter(case_ == T),
           aes(y = ndvi), alpha = 0.25, length = unit(0.015, "npc"), sides="l") +
  scale_y_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  theme_bw() 

ggsave(paste0(plot_save_path, "/ndvi_tensor_gam_response_id", which_buffalo, ".png"),
       width = 150, height = 80, units = "mm", dpi = 600)

```

### Pull out a response curve from one of the hours

```{r}

# Get predictions on log-scale with standard errors
predictions <- predict(ssf_gam2d_temporal2, 
                      newdata = pred_data_hour9_17, 
                      type = "link",  # log-scale
                      se.fit = TRUE)

pred_data_hour9_17 <- pred_data_hour9_17 %>% 
  mutate(
    # Extract log RSS and standard errors
    log_rss = predictions$fit,
    se_log_rss = predictions$se.fit,
    # Calculate confidence intervals on log-scale
    log_rss_lower = log_rss - 1.96 * se_log_rss,
    log_rss_upper = log_rss + 1.96 * se_log_rss,
    # Transform to RSS scale
    rss = exp(log_rss),
    rss_lower = exp(log_rss_lower),
    rss_upper = exp(log_rss_upper)
  )

# plot the response curve
ggplot(data = pred_data_hour9_17) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey") +
  geom_ribbon(aes(x = ndvi, 
                  ymin = rss_lower, ymax = rss_upper,
                  fill = as.factor(hour)), 
              alpha = 0.5) +
  geom_line(aes(x = ndvi, y = rss, colour = as.factor(hour)), 
            linewidth = 1) +
  geom_rug(data = buffalo_ssf_data %>% filter(case_ == T), 
           aes(x = ndvi), alpha = 0.25, sides="b") +
  scale_colour_manual(values = c("orange", "skyblue")) +
  scale_fill_manual(values = c("orange", "skyblue")) +
  scale_x_continuous(breaks = seq(-0.2, 0.8, by = 0.1)) +
  scale_y_continuous(limits = c(0, 5.5)) +
  labs(x = "NDVI", y = "RSS") +
  theme_classic() +
  theme(legend.position = "none")

ggsave(paste0(plot_save_path, "/ndvi_gam2d_response_sample_hour_id", which_buffalo, ".png"),
       width = 70, height = 80, units = "mm", dpi = 600)

```

# References

::: {#refs}
:::

